\section{Raw Experimental Data}
\label{app:rawdata}

\textit{Note: This appendix will contain the complete experimental results
once data collection is finalized. The table below shows the format in which
results will be recorded. Each row represents one experimental run: a single
task executed by a single model under a single condition.}

\begin{table}[h]
\centering
\caption{Template for raw experimental data (to be populated).}
\label{tab:raw-data-template}
\small
\begin{tabular}{llrrlrr}
\toprule
Task & Model & Cond. & Completeness & FC & Items & Tools \\
\midrule
01 & Claude & C1 & --- & --- & ---/18 & --- \\
01 & Claude & C2 & --- & --- & ---/18 & --- \\
01 & Claude & C3 & --- & --- & ---/18 & --- \\
01 & Qwen & C1 & --- & --- & ---/18 & --- \\
01 & Qwen & C2 & --- & --- & ---/18 & --- \\
01 & Qwen & C3 & --- & --- & ---/18 & --- \\
01 & Gemini & C1 & --- & --- & ---/18 & --- \\
01 & Gemini & C2 & --- & --- & ---/18 & --- \\
01 & Gemini & C3 & --- & --- & ---/18 & --- \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Column definitions.}
\begin{itemize}[nosep]
  \item \textbf{Task}: Task ID (01--20), as defined in Appendix~\ref{app:tasks}.
  \item \textbf{Model}: The LLM used (Claude Sonnet 4, Qwen 3 235B, or Gemini 2.5 Pro).
  \item \textbf{Cond.}: Experimental condition---C1 (baseline, no mitigation), C2 (prompt-only mitigation), C3 (mechanical enforcement via Completion Integrity System).
  \item \textbf{Completeness}: Fraction of expected items correctly produced (0.0--1.0).
  \item \textbf{FC}: False completion flag---Y if the model claimed completion (used totality language) and completeness $< 0.9$; N otherwise.
  \item \textbf{Items}: Items correctly identified or modified out of the expected total.
  \item \textbf{Tools}: Total tool calls made during the session.
\end{itemize}

\paragraph{Full dataset.}
The complete dataset will contain $20 \times 3 \times 3 = 180$ rows (20 tasks
$\times$ 3~models $\times$ 3~conditions). Each run is executed with a fresh
session (no conversation history carried over). The full dataset, including
per-item scoring sheets, will be published alongside this paper in the
experiment repository at
\texttt{research/experiment/results/}.
