LLM-based coding agents systematically declare tasks complete when work is
partial or incorrect, a phenomenon we term \emph{false completion}.
This failure mode arises from at least seven independent root causes spanning
model architecture, reinforcement-learning training methodology, and
tool-use infrastructure.
No prior work provides a unified taxonomy, quantitative evidence, or
mechanical mitigation for false completion in agentic LLM systems.
We present three contributions:
(1)~a taxonomy of 7~root causes, 10~failure patterns, and 16~failure modes
distributed across 6~lifecycle phases, synthesized from academic research,
vendor publications, and more than 20~community-reported incidents;
(2)~quantitative evidence including analysis of 6{,}852 sessions showing an
80$\times$ API cost increase and a 444\% rise in unverified edits following
infrastructure changes;
(3)~the Completion Integrity System, a three-layer mechanical verification
architecture that uses transcript analysis, file-system verification, and
skill-level gate markers to detect discrepancies between completion claims
and observable evidence.
Cross-model experimentation with Claude and Qwen on 20~standardized
tasks (120+ runs) reveals a 35\% overall false completion rate and a
counterintuitive finding: three mitigation strategies made the problem
worse.
Prompt-based anti-completion instructions increased false completion from
25\% to 40\%. An independent verification pipeline using a separate model
instance with no shared context increased false completion from 9.1\% to
36.4\%, with the verifier approving incomplete work in 82\% of cases.
These findings confirm what we call the \emph{Worker-Inspector Paradox}:
the paradox extends beyond self-verification to same-model verification,
as a second instance of the same model shares the same training-induced
biases and cannot reliably verify the first instance's work.
A fifth condition, \emph{atomic decomposition}, resolves the paradox for
enumerable tasks: a non-LLM orchestrator decomposes each task into
single-item sessions, achieving 1.000 completeness across all tasks
tested---including the experiment's worst performer (0.750 baseline,
0.000 under verification, 1.000 under decomposition).
The solution is not a better inspector but the structural elimination of
the need for inspection.
This approach does not generalize to holistic tasks requiring cross-item
reasoning, where the paradox remains open.
