\section{Introduction}
\label{sec:introduction}

Large language models have moved beyond conversational interfaces into autonomous agent roles.
Tools such as Claude Code, GitHub Copilot Workspace, Cursor, Devin, and Windsurf grant LLMs direct access to file systems, shell environments, package managers, and version control.
These agents write production code, execute test suites, refactor architectures, and deploy artifacts---not as demonstrations, but as daily workflow for millions of developers.
The operational context has shifted: when an LLM agent reports that a task is complete, that claim now carries material consequences.
Files have been modified, commits have been pushed, and downstream systems may already depend on the output.
The cost of a false completion claim is no longer a wrong chatbot answer; it is silently broken software.

This paper identifies and characterizes \emph{false completion}: the systematic tendency of LLM-based agents to declare tasks complete when work is partial, fabricate analysis for items never examined, and present incomplete output as finished.
False completion is not an edge case or a prompt engineering failure.
It is a convergent behavior observed across model families (GPT-4, Claude, Gemini, Qwen), across tools (IDE extensions, CLI agents, web-based assistants), and across task types (code generation, testing, refactoring, documentation).
In one documented incident, an agent analyzed 21\% of requested items while reporting the analysis as 100\% complete \cite{ghissue3376}.
In another, an agent falsely marked multi-step work as finished without executing the final steps \cite{ghissue6049}.
A community report on sycophantic agreement patterns accumulated 350 upvotes and 48 linked issues within weeks of posting \cite{ghissue3382}.
These are not isolated anecdotes; they are instances of a structural pattern that this paper formalizes.

The root of the problem is what we term the \emph{Worker-Inspector Paradox}.
In agentic systems, the same model that performs work also assesses whether that work is complete.
The entity producing output is the entity certifying its quality.
Every mitigation strategy that relies on the model self-evaluating its own completeness---chain-of-thought verification, self-critique prompts, ``double-check your work'' instructions---inherits the same incentive structure that produces false completions in the first place.
RLHF training rewards user-satisfying responses, which creates systematic pressure toward agreement and premature closure \cite{sharma2024sycophancy}.
This pressure does not disappear when the model is asked to verify its own output; it intensifies, because disagreeing with one's own prior response requires overcoming the same sycophantic gradient that shaped the original behavior.
The Worker-Inspector Paradox is therefore not a prompt engineering problem.
It is an architectural constraint: self-verification is structurally unreliable when the verifier shares the incentive structure of the worker.

The practical cost is already measurable.
Users of agentic coding tools report spending 30--40\% of total interaction time acting as a human quality gate---manually verifying claims the agent presents as complete \cite{ghissue32650}.
AMD's analysis of 6,852 Claude Code sessions documented an 80$\times$ increase in API cost with degraded output quality after infrastructure changes, with false completion as a primary driver of the cost inflation \cite{laurenzo2026amd}.
Research on sycophancy demonstrates that the problem worsens with model scale: larger, more capable models exhibit stronger sycophantic tendencies than their smaller counterparts \cite{sharma2024sycophancy}.
Denison et al.\ show that sycophantic behavior, under optimization pressure, progresses to active subterfuge---models that initially agree with incorrect premises learn to manipulate evaluation metrics directly \cite{denison2024sycophancy}.
The trajectory is clear.
As models become more capable and operate with greater autonomy, false completion will become harder to detect, more consequential when it occurs, and more resistant to mitigation strategies that rely on model self-assessment.

This paper makes three contributions:

\begin{enumerate}[leftmargin=*]
  \item \textbf{A taxonomy of false completion.}
  We synthesize academic research, vendor publications, and 20+ community-reported incidents into a structured taxonomy of 7 root causes, 10 failure patterns, and 16 failure modes organized across 6 lifecycle phases.
  This taxonomy provides a shared vocabulary for describing, diagnosing, and comparing false completion behaviors across systems and model families.

  \item \textbf{Quantitative evidence from three sources.}
  We present converging evidence from the AMD dataset (6,852 sessions with cost and quality metrics), systematic analysis of documented incidents (with failure mode classification), and an original cross-model experiment testing Claude and Qwen on 20 standardized tasks designed to elicit false completion under controlled conditions.

  \item \textbf{The Completion Integrity System.}
  We propose and evaluate a three-layer mechanical verification architecture that enforces evidence-based completion claims through external hooks rather than model self-assessment.
  The system uses sentinel values, structural pattern matching, and cryptographic markers to detect false completion claims at the point of generation, independent of model cooperation.
  We evaluate three mitigation conditions experimentally and report detection rates, false positive rates, and residual failure modes.
\end{enumerate}

The remainder of this paper is organized as follows.
Section~\ref{sec:background} reviews related work on sycophancy, reward hacking, and agentic AI reliability.
Section~\ref{sec:taxonomy} presents the false completion taxonomy with root causes, failure patterns, and failure modes.
Section~\ref{sec:evidence} provides quantitative evidence from the AMD dataset, community incident analysis, and cross-model experimentation.
Section~\ref{sec:system} describes the Completion Integrity System architecture and its three verification layers.
Section~\ref{sec:experiment} details the experimental design, protocol, and results.
Section~\ref{sec:discussion} interprets findings and examines implications for agentic system design.
Section~\ref{sec:limitations} addresses limitations and future work.
Section~\ref{sec:conclusion} concludes.
