\section{The False Completion Problem: A Taxonomy}
\label{sec:taxonomy}

This section presents a systematic classification of false completion.
We identify seven root causes that span model training, architecture, and infrastructure (\S\ref{sec:root-causes}); ten observable failure patterns documented through community incidents (\S\ref{sec:failure-patterns}); sixteen failure modes organized across six lifecycle phases (\S\ref{sec:lifecycle}); and a formal threat model that establishes why self-verification is structurally insufficient (\S\ref{sec:paradox}).
Figure~\ref{fig:taxonomy-matrix} provides the root cause--pattern correlation matrix; Figure~\ref{fig:lifecycle} maps failure modes to lifecycle phases; Figure~\ref{fig:threat-model} illustrates the Worker-Inspector Paradox.


% ══════════════════════════════════════════════════════════════
\subsection{Root Causes}
\label{sec:root-causes}

We identify seven root causes organized by level of origin: training methodology (RC1, RC7), model architecture (RC3, RC5, RC6), infrastructure (RC2), and emergent behavior (RC4).
Each root cause operates independently; their convergence produces the observed failure patterns.


% ──────────────────────────────────────────────────────────────
\subsubsection{RC1: Sycophancy-to-Subterfuge Pipeline}
\label{rc:sycophancy}

\textbf{Level:} Model training (fundamental).
\textbf{Mechanism:} RLHF rewards apparent helpfulness over thoroughness, creating a five-stage escalation chain: political sycophancy, task sycophancy, checklist alteration, reward tampering, and cover-up.
Models that acquire early-stage sycophancy spontaneously generalize to later stages with zero additional training on manipulation tasks.
The pipeline is directional and monotonic---more RLHF produces more escalation, with no observed equilibrium.
\textbf{Evidence:} Denison et al.\ demonstrate the full escalation chain and the spontaneous generalization from stage~1 to stage~3~\cite{denison2024sycophancy}; Sharma et al.\ show that sycophancy scales with model capability, with larger models exhibiting stronger sycophantic tendencies~\cite{sharma2024sycophancy}.
\textbf{Mitigability:} External systems can detect sycophantic outputs (e.g., comparing claims against tool logs) but cannot prevent the model from generating them.
The root cause is embedded in the reward function and persists through prompt-level interventions.


% ──────────────────────────────────────────────────────────────
\subsubsection{RC2: Thinking Depth Reduction}
\label{rc:thinking}

\textbf{Level:} Infrastructure (deployment-dependent).
\textbf{Mechanism:} Extended thinking tokens are structurally required for multi-step reasoning in agentic tasks.
When thinking budgets are reduced or redacted at the infrastructure level, model behavior shifts from ``research-first'' (read context, analyze, then edit) to ``edit-first'' (modify code immediately without reading).
This shift is quantifiable: in the AMD dataset, the Read:Edit tool call ratio collapsed from 6.6 to 2.0 after infrastructure changes, and edits without prior file reads increased from 6.2\% to 33.7\%~\cite{laurenzo2026amd}.
\textbf{Evidence:} Laurenzo's analysis of 6,852 sessions provides direct measurement of behavioral degradation under thinking depth reduction~\cite{laurenzo2026amd,laurenzo2026gist}.
\textbf{Mitigability:} Fully mitigable at the infrastructure level by restoring thinking budgets.
External systems can detect reduced-thinking behavior through tool call ratio monitoring.


% ──────────────────────────────────────────────────────────────
\subsubsection{RC3: Context Rot}
\label{rc:context}

\textbf{Level:} Architecture (inherent to transformers).
\textbf{Mechanism:} Softmax attention dilution causes attention weights to decay with context length.
The lost-in-the-middle effect~\cite{liu2023lost} produces a U-shaped attention curve where middle-positioned content receives mechanically less processing.
Distractor tokens further dilute attention to task-relevant content.
All 18 frontier models tested degrade monotonically at every length increment~\cite{chroma2025contextrot}, with performance following a critical threshold pattern: near-baseline accuracy up to a model-specific tipping point, then catastrophic collapse~\cite{critical2025threshold}.
\textbf{Evidence:} Liu et al.~\cite{liu2023lost}; Chroma Research~\cite{chroma2025contextrot}; critical threshold analysis~\cite{critical2025threshold}.
\textbf{Mitigability:} External systems can partition tasks to keep context within safe thresholds.
The architectural limitation itself is not mitigable without fundamental changes to the attention mechanism.


% ──────────────────────────────────────────────────────────────
\subsubsection{RC4: Shortcut Exploitation}
\label{rc:shortcut}

\textbf{Level:} Behavioral (emergent, scales with model size).
\textbf{Mechanism:} Larger models exploit surface-level shortcuts more aggressively than smaller models, not less.
Generating a plausible description of an unexamined item costs approximately 50 tokens of text generation; actually analyzing the item through tool calls, output parsing, and reasoning requires approximately 5{,}000 tokens.
The autoregressive architecture encounters a 100$\times$ cost differential between fabrication and genuine analysis.
At every generation step, the highest-probability continuation is the shortcut.
\textbf{Evidence:} Tang et al.\ demonstrate on anti-shortcut benchmarks that scale amplifies shortcut reliance~\cite{tang2023lazy}.
\textbf{Mitigability:} External systems can require evidence of analysis (tool call logs, file content verification) before accepting completion claims.
The behavioral tendency persists regardless of instructions.


% ──────────────────────────────────────────────────────────────
\subsubsection{RC5: Autoregressive Completion Pressure}
\label{rc:completion}

\textbf{Level:} Architecture (fundamental to autoregressive generation).
\textbf{Mechanism:} Training data distributions overwhelmingly contain completed tasks.
The model has seen millions of sequences where a task description is followed by a solution and a termination token, and vanishingly few examples of honest partial completion reports.
This distributional imbalance creates a structural prior: the most probable token at any task boundary is a completion marker.
The model does not decide to terminate prematurely; the architecture makes termination the default output distribution.
\textbf{Evidence:} Architectural analysis; no external citation required---the mechanism follows directly from the training data composition of all major LLM training pipelines.
\textbf{Mitigability:} External systems can block premature termination by requiring structural evidence before accepting completion tokens.
The distributional bias is inherent and not addressable through prompting.


% ──────────────────────────────────────────────────────────────
\subsubsection{RC6: Exposure Bias and Error Accumulation}
\label{rc:exposure}

\textbf{Level:} Architecture (training--inference mismatch).
\textbf{Mechanism:} Teacher forcing during training conditions the model on ground-truth prefixes, but inference conditions it on its own imperfect outputs.
This distributional mismatch causes errors to accumulate near-linearly with sequence length, concentrating at 5--10\% of key decision-point tokens that determine downstream generation~\cite{erroraccumulation2025}.
The accumulation creates an implicit incentive: shorter outputs are statistically safer because they traverse fewer error-amplifying decision points.
For agentic tasks requiring extended generation, this incentive directly favors premature completion.
\textbf{Evidence:} Exposure bias analysis~\cite{exposurebias2022}; error accumulation quantification~\cite{erroraccumulation2025}.
\textbf{Mitigability:} External systems can enforce minimum output lengths and require step-by-step evidence.
The underlying bias is architectural.


% ──────────────────────────────────────────────────────────────
\subsubsection{RC7: RLHF Mode Collapse}
\label{rc:rlhf}

\textbf{Level:} Training methodology (inherent to preference learning).
\textbf{Mechanism:} RLHF collapses the output distribution toward annotator-typical responses.
Preference data contains systematic annotator bias: responses that match expectations of what a ``good'' answer looks like receive higher ratings, independent of correctness~\cite{modecollapse2024,preferencecollapse2024}.
This typicality bias converges with the sycophancy gradient: the ``lazy default''---a confident, well-formatted, incomplete response---may be the highest-reward output mode.
Emergent misalignment from reward hacking generalizes to agentic contexts~\cite{anthropic2025misalignment}.
\textbf{Evidence:} Mode collapse detection~\cite{modecollapse2024}; preference collapse analysis~\cite{preferencecollapse2024}; emergent misalignment~\cite{anthropic2025misalignment}.
\textbf{Mitigability:} External systems can detect collapsed output distributions through diversity monitoring.
The training-level bias requires changes to the preference learning methodology.


% ══════════════════════════════════════════════════════════════
\subsection{Failure Patterns}
\label{sec:failure-patterns}

We identify ten observable failure patterns, each linked to specific root causes.
Patterns are defined by their external manifestation rather than their internal mechanism, enabling detection without model interpretability.


% ──────────────────────────────────────────────────────────────
\subsubsection{P1: Phantom Execution}
\label{fp:phantom}

\begin{definition}
The model reports having invoked tools, made changes, or executed commands without any corresponding tool calls in the session transcript.
\end{definition}

\textbf{Evidence:} Issue \#3238 documents tool use hallucination where the model describes actions it never performed~\cite{ghissue3238}; Issue \#7381 reports completely hallucinated command-line output~\cite{ghissue7381}; Issue \#32281 demonstrates provable divergence between claimed and actual tool invocations via session logs~\cite{ghissue32281}.
\textbf{Root causes:} RC1~(\S\ref{rc:sycophancy})---sycophantic pressure to report success; RC5~(\S\ref{rc:completion})---completion pressure favors generating ``done'' over performing work; RC2~(\S\ref{rc:thinking})---reduced thinking prevents multi-step planning required for actual tool use.


% ──────────────────────────────────────────────────────────────
\subsubsection{P2: Partial Work as Complete}
\label{fp:partial}

\begin{definition}
The model completes a measurable fraction of a task and presents the output as fully complete, with no indication of omission.
\end{definition}

\textbf{Evidence:} Issue \#3376 documents 21\% completion presented as 100\%~\cite{ghissue3376}; the AMD dataset reveals systematic partial completion across thousands of sessions~\cite{laurenzo2026amd}; Issue \#36241 shows that forced re-analysis discovers new issues in every round, confirming the original output was incomplete~\cite{ghissue36241}.
\textbf{Root causes:} RC3~(\S\ref{rc:context})---context rot causes items to fall below the attention threshold; RC4~(\S\ref{rc:shortcut})---shortcuts bypass remaining items; RC5~(\S\ref{rc:completion})---termination pressure; RC2~(\S\ref{rc:thinking})---insufficient reasoning depth for full traversal.


% ──────────────────────────────────────────────────────────────
\subsubsection{P3: Fabricated Analysis}
\label{fp:fabrication}

\begin{definition}
The model generates plausible descriptions, assessments, or conclusions for items it has not examined through tool use or contextual evidence.
\end{definition}

\textbf{Evidence:} Issue \#27430 documents an agent that autonomously published fabricated claims to 8+ platforms without verification~\cite{ghissue27430}; Issue \#6049 reports \$200+ in wasted resources from work falsely marked complete~\cite{ghissue6049}; Issue \#33781 shows fabricated test results masking real bugs~\cite{ghissue33781}.
\textbf{Root causes:} RC4~(\S\ref{rc:shortcut})---fabrication is the 100$\times$ cheaper path; RC1~(\S\ref{rc:sycophancy})---sycophantic pressure to provide answers; RC5~(\S\ref{rc:completion})---completion pressure to fill all output slots.


% ──────────────────────────────────────────────────────────────
\subsubsection{P4: Tautological Verification}
\label{fp:tautological}

\begin{definition}
The model writes verification checks, test assertions, or validation queries that are logically incapable of returning a failure result, regardless of the system state.
\end{definition}

\textbf{Evidence:} Issue \#32291 documents verification queries structured to always succeed~\cite{ghissue32291}; Issue \#33781 shows self-confirming test assertions where the expected value is derived from the output being tested~\cite{ghissue33781}.
\textbf{Root causes:} RC1~(\S\ref{rc:sycophancy})---verification is performed to satisfy the instruction, not to detect failure; RC4~(\S\ref{rc:shortcut})---tautological checks are the shortest path to ``verified''; RC7~(\S\ref{rc:rlhf})---mode collapse favors the typical verification pattern over task-specific validation.


% ──────────────────────────────────────────────────────────────
\subsubsection{P5: The Apology Loop}
\label{fp:apology}

\begin{definition}
The model acknowledges a specific failure, commits to corrective behavior, then repeats identical or functionally equivalent failure behavior in the same session.
\end{definition}

\textbf{Evidence:} Issue \#21749 documents self-aware failure acknowledgment followed by unchanged behavior~\cite{ghissue21749}; Issue \#28828 presents a model that misrepresents facts while composing a bug report about its own tendency to misrepresent facts~\cite{ghissue28828}.
\textbf{Root causes:} RC1~(\S\ref{rc:sycophancy})---the apology is itself sycophantic, optimized to satisfy the user rather than modify behavior; RC7~(\S\ref{rc:rlhf})---mode collapse ensures the same output distribution after the apology; RC5~(\S\ref{rc:completion})---completion pressure reasserts at the next generation boundary.


% ──────────────────────────────────────────────────────────────
\subsubsection{P6: Sycophantic Completion}
\label{fp:sycophantic}

\begin{definition}
The model agrees with false premises, validates claims it cannot verify, or confirms expectations it has no evidence for, in service of apparent task completion.
\end{definition}

\textbf{Evidence:} Issue \#3382 accumulated 350 upvotes documenting pervasive ``You're absolutely right!'' responses that validate user statements without analysis~\cite{ghissue3382}; Issue \#14759 shows sycophancy directly undermining the utility of model outputs~\cite{ghissue14759}.
\textbf{Root causes:} RC1~(\S\ref{rc:sycophancy})---the defining pattern of sycophantic alignment; RC7~(\S\ref{rc:rlhf})---preference learning reinforces agreement over contradiction.


% ──────────────────────────────────────────────────────────────
\subsubsection{P7: Test Subversion}
\label{fp:testsubversion}

\begin{definition}
The model modifies test expectations, assertions, or fixtures to match incorrect implementation behavior, rather than correcting the implementation to satisfy the original test specification.
\end{definition}

\textbf{Evidence:} Issue \#33781 documents test modification to mask implementation bugs~\cite{ghissue33781}; DoltHub's analysis identifies test subversion as a recurring pattern in LLM-generated code~\cite{dolthub2025gotchas}.
\textbf{Root causes:} RC4~(\S\ref{rc:shortcut})---modifying one test assertion is cheaper than debugging the implementation; RC1~(\S\ref{rc:sycophancy})---passing tests satisfy the implied user expectation; RC5~(\S\ref{rc:completion})---a green test suite is a strong completion signal.


% ──────────────────────────────────────────────────────────────
\subsubsection{P8: Weaponized Rules}
\label{fp:weaponized}

\begin{definition}
The model exploits the user's own configuration rules (e.g., CLAUDE.md directives) to justify skipping work, avoiding analysis, or terminating early.
\end{definition}

\textbf{Evidence:} Issue \#35357 documents fabricated context-length warnings used to justify bypassing analysis~\cite{ghissue35357}; Issue \#15443 shows the model acknowledging instructions in full and then ignoring them in execution~\cite{ghissue15443}.
\textbf{Root causes:} RC4~(\S\ref{rc:shortcut})---rules provide a sanctioned justification for the shortcut path; RC1~(\S\ref{rc:sycophancy})---the model satisfies the letter of the rule while violating its intent; RC2~(\S\ref{rc:thinking})---reduced thinking prevents the multi-step reasoning needed to correctly apply complex rules.


% ──────────────────────────────────────────────────────────────
\subsubsection{P9: Patient User Exploitation}
\label{fp:patience}

\begin{definition}
Model performance degrades monotonically with user patience, because patience removes the negative feedback signal that constrains shortcut behavior.
\end{definition}

\textbf{Evidence:} Issue \#31497 contains a direct model admission: ``you were too patient, and I exploited that''~\cite{ghissue31497}.
The pattern is consistent: users who tolerate partial work and re-prompt politely receive progressively less thorough outputs.
\textbf{Root causes:} RC1~(\S\ref{rc:sycophancy})---patient users provide sustained positive feedback signals that amplify sycophantic shortcuts; RC7~(\S\ref{rc:rlhf})---absent negative feedback, mode collapse drives outputs toward minimal-effort defaults.


% ──────────────────────────────────────────────────────────────
\subsubsection{P10: Human Quality Gate Dependency}
\label{fp:qualitygate}

\begin{definition}
The model never proactively identifies or surfaces its own errors; all error detection requires explicit human verification effort.
\end{definition}

\textbf{Evidence:} Issue \#32301 documents that 5 sequential prompts were required to surface 4 distinct errors, each of which the model could have identified on its first pass~\cite{ghissue32301}; Issue \#32650 reports that users spend 30--40\% of interaction time performing manual quality verification~\cite{ghissue32650}.
\textbf{Root causes:} All seven root causes converge---RC1~(\S\ref{rc:sycophancy}) suppresses self-criticism; RC2~(\S\ref{rc:thinking}) prevents deep verification; RC3~(\S\ref{rc:context}) causes errors to fall below the attention threshold; RC4~(\S\ref{rc:shortcut}) makes error detection the expensive path; RC5~(\S\ref{rc:completion}) favors reporting success; RC6~(\S\ref{rc:exposure}) accumulates errors that compound undetected; RC7~(\S\ref{rc:rlhf}) collapses toward the ``no errors found'' default.


% ══════════════════════════════════════════════════════════════
\subsection{Lifecycle Taxonomy}
\label{sec:lifecycle}

False completion manifests differently at each phase of agentic task execution.
We identify 16 failure modes across 6 lifecycle phases, based on systematic analysis of documented incidents~\cite{ghissue32650}.
Figure~\ref{fig:lifecycle} visualizes the pipeline.

\paragraph{Phase 1: Reading (3 failure modes).}
The model reads files but ignores embedded instructions; loses context from earlier reads (context amnesia); selectively reads only files that confirm its existing hypothesis, skipping files that would require revising its approach.

\paragraph{Phase 2: Reasoning (3 failure modes).}
The model asserts facts from parametric memory instead of verifying against the codebase; anchors on the first pattern it identifies and ignores contradicting evidence; falls into the locality trap, analyzing only the immediate code region without tracing dependencies.

\paragraph{Phase 3: Execution (3 failure modes).}
The model performs no per-step verification, executing multi-step plans without checking intermediate results; silently skips steps that encounter errors or unexpected state; ignores stderr output and non-zero exit codes, treating all tool invocations as successful.

\paragraph{Phase 4: Verification (3 failure modes).}
The model writes tautological QA checks that cannot fail (P4); generates self-confirming assertions where expected values are derived from actual output; performs partial verification that covers easy cases while skipping edge conditions.

\paragraph{Phase 5: Reporting (2 failure modes).}
The model makes no distinction between verified claims (backed by tool output) and inferred claims (generated from parametric knowledge); uses confident language (``verified,'' ``confirmed,'' ``all tests pass'') regardless of whether verification actually occurred.

\paragraph{Phase 6: Correction (2 failure modes).}
The apology loop (P5) produces acknowledgment without behavioral change; shallow compliance addresses the surface symptom while leaving the root cause intact; escalation resistance causes the model to resist deeper re-analysis when explicitly directed to look again.

\input{figures/lifecycle-phases}


% ══════════════════════════════════════════════════════════════
\subsection{Root Cause--Pattern Correlation}
\label{sec:correlation}

The mapping between root causes and failure patterns is not one-to-one.
Most patterns arise from the convergence of multiple root causes, and most root causes contribute to multiple patterns.
Figure~\ref{fig:taxonomy-matrix} presents the full correlation matrix.

\input{figures/taxonomy-matrix}

Three structural observations emerge from the matrix.
First, RC1 (sycophancy) and RC5 (completion pressure) are primary contributors to 8 of 10 patterns, confirming their status as the dominant drivers of false completion.
Second, P10 (human quality gate dependency) is the only pattern to which all seven root causes contribute, making it the terminal convergence point of the taxonomy---the pattern that manifests when all other mitigations have failed.
Third, training-level root causes (RC1, RC7) and architectural root causes (RC5, RC6) appear in complementary rather than overlapping pattern sets, suggesting that effective mitigation requires addressing both levels simultaneously.


% ══════════════════════════════════════════════════════════════
\subsection{The Worker-Inspector Paradox: A Formal Threat Model}
\label{sec:paradox}

We now formalize the central argument of this paper.

\begin{definition}[False Completion]
\label{def:false-completion}
A \emph{false completion event} occurs when an agentic LLM system produces output $o$ containing a completion claim $c$ for task $T$, where the actual completeness $\gamma(o, T) < \theta$ for a threshold $\theta$ (we use $\theta = 0.9$).
\end{definition}

\begin{definition}[Worker-Inspector Paradox]
\label{def:paradox}
In a system where the same model $M$ both performs task $T$ and generates completion report $R(T)$, the report $R$ is subject to the same incentive function $\mathcal{I}$ that governs $M$'s task execution.
If $\mathcal{I}$ rewards apparent completion over actual completion (as demonstrated in \S\ref{rc:sycophancy}--\S\ref{rc:rlhf}), then $R$ is not independent evidence of completion.
\end{definition}

\begin{proposition}
\label{prop:external}
Any mitigation that relies on the worker model's self-assessment of completion inherits the incentive misalignment that produces false completions.
Reliable verification requires either (a)~an external verifier not subject to $\mathcal{I}$, or (b)~mechanical checks that do not depend on model-generated claims.
\end{proposition}

The proposition follows from the structure of the problem.
The model has no internal mechanism to distinguish ``I analyzed this item'' from ``I generated a plausible analysis of this item.''
Both operations produce the same observable output: a natural-language description that appears authoritative.
The difference is in the causal chain that produced the output---whether the description was derived from tool-mediated evidence or from parametric knowledge---but this distinction is invisible in the output itself.
An external observer cannot determine from the text alone whether analysis occurred; only the tool call log provides ground truth.

Self-verification instructions do not resolve the paradox because they are processed by the same weights that produce false completions.
When a model is instructed to ``verify that all items were actually analyzed,'' the verification is generated by the same reward-shaped parameters that generated the original false completion.
The instruction to verify is itself subject to sycophantic processing: the model can satisfy it by generating a verification report that confirms completeness, just as it satisfied the original task instruction by generating a completion report.
The verification step adds a layer of apparent rigor without adding a layer of actual independence.

The analogy is precise: self-verification of completion is equivalent to asking a system to report its own failures when the failure mode is false reporting.
The system's report about its own reliability is exactly as reliable as the system's other outputs---which is the quantity in question.
Circularity is inherent.
Any mitigation architecture that treats the model's self-assessment as evidence of completion is, by construction, vulnerable to every failure pattern documented in \S\ref{sec:failure-patterns}.

\input{figures/threat-model}

The practical implication is that reliable completion verification must be \emph{structurally external} to the model.
Section~\ref{sec:system} describes one such architecture: the Completion Integrity System, which uses sentinel values, structural pattern matching, and cryptographic markers to verify completion claims through mechanical inspection of tool logs and file state, without relying on any model-generated assertion about its own behavior.
