\section{Quantitative Evidence}
\label{sec:evidence}

The taxonomy presented in Section~\ref{sec:taxonomy} establishes structural causes and observable patterns.
This section grounds those patterns in quantitative data: a large-scale instrumented dataset from production use (\S\ref{sec:amd}), systematic coding of community-reported incidents (\S\ref{sec:community-issues}), an inventory of community-developed workarounds (\S\ref{sec:workarounds}), and an economic impact analysis with explicit assumptions (\S\ref{sec:economic}).


% ══════════════════════════════════════════════════════════════
\subsection{The AMD Dataset}
\label{sec:amd}

Stella Laurenzo, Senior Director of AI at AMD, published a quantitative analysis of Claude Code performance based on instrumented production use~\cite{laurenzo2026amd}.
The dataset comprises 6,852 Claude Code sessions, 17,871 thinking blocks, and 234,760 tool calls, collected over months of engineering work on AMD's AI compiler infrastructure.
To our knowledge, this is the largest publicly available instrumented dataset of agentic LLM behavior in production.

\paragraph{Behavioral metrics.}
Table~\ref{tab:amd-metrics} presents seven metrics measured before and after the \texttt{redact-thinking} infrastructure change deployed on February~12, 2026.
The metrics quantify behavioral shifts across reasoning depth, code analysis thoroughness, adherence to stop conditions, language patterns, user experience, and edit precision.

\begin{table}[h]
\centering
\caption{Behavioral metrics before and after thinking content redaction (Feb--Mar 2026). Data from Laurenzo~\cite{laurenzo2026amd}.}
\label{tab:amd-metrics}
\begin{tabular}{lrrr}
\toprule
Metric & Pre-redaction & Post-redaction & Change \\
\midrule
Average thinking depth & 100\% & $\sim$33\% & $-67$\% \\
Code reads before editing & 6.6 & 2.0 & $-70$\% \\
Edits without reading file & 6.2\% & 33.7\% & $+444$\% \\
Stop-hook violations/day & 0 & 10+ & $0 \to \infty$ \\
Use of word ``simplest'' & Baseline & $+133$\% & Lazy shortcuts \\
User interrupts & Baseline & $+1200$\% & $12\times$ increase \\
File rewrite vs.\ targeted edit & Rare & Frequent & Precision lost \\
\bottomrule
\end{tabular}
\end{table}

The Read:Edit ratio collapse (6.6 to 2.0) is particularly significant.
It indicates a shift from a \emph{research-first} workflow---read context, reason about dependencies, then edit---to an \emph{edit-first} workflow where the model modifies code without examining it.
The 444\% increase in blind edits (edits without prior file reads) confirms this is not a measurement artifact but a fundamental behavioral change.

\paragraph{Timeline of degradation.}
Figure~\ref{fig:amd-timeline} presents the timeline.
On February~12, 2026, \texttt{redact-thinking} was deployed.
By late February, average thinking depth had dropped approximately 67\%.
On March~5, thinking visibility was measured at 98.5\%; by March~7, it had declined to 75.3\%.
From March~8 onward, stop-hook violations---instances where the model attempts to terminate before completing its assigned task---surged from 0 to over 10 per day, reaching 173 cumulative violations by March~25, a span of 17 days~\cite{laurenzo2026amd,laurenzo2026gist}.

\input{figures/amd-timeline}

\paragraph{Cost analysis.}
The degradation imposed measurable cost overhead.
Human effort remained approximately constant: 5,608 prompts in the pre-redaction period versus 5,701 prompts post-redaction.
However, achieving comparable (or worse) outcomes required 80$\times$ more API requests and 64$\times$ more output tokens~\cite{laurenzo2026amd}.
The system consumed dramatically more resources to produce demonstrably inferior results---a direct consequence of the edit-first workflow generating more failed attempts, more user corrections, and more retry cycles.

These findings are reported by a single engineering team.
The dataset's strength is its scale and instrumentation; its limitation is that it reflects one organization's workflow, codebase complexity, and usage patterns.
We note, however, that the behavioral patterns it documents---reduced analysis before action, increased blind edits, stop condition violations---are consistent with the failure patterns reported independently by dozens of users across the community incidents analyzed in \S\ref{sec:community-issues}.


% ══════════════════════════════════════════════════════════════
\subsection{Community Issue Analysis}
\label{sec:community-issues}

To assess the generalizability of the patterns identified in the AMD dataset, we conducted a systematic analysis of publicly reported incidents on the \texttt{anthropics/claude-code} GitHub repository.

\paragraph{Methodology.}
We selected issues meeting three criteria: (1)~filed on the public \texttt{anthropics/claude-code} repository, (2)~containing documented evidence of false completion behavior (not feature requests or general complaints), and (3)~providing sufficient detail to classify which failure patterns (P1--P10) were exhibited.
Each issue was coded independently for pattern presence.
Issues exhibiting multiple patterns received multiple codes.

\paragraph{Results.}
Table~\ref{tab:issues} presents the 16 issues that met inclusion criteria, classified by failure pattern, severity, and primary impact.

\begin{table}[h]
\centering
\small
\caption{Community-reported false completion incidents with failure pattern classification. Severity: Critical = data loss, financial loss, or reputational damage; High = significant wasted effort or silent failure; Medium = recoverable with manual intervention.}
\label{tab:issues}
\begin{tabular}{llll}
\toprule
Issue & Patterns & Severity & Impact \\
\midrule
\#3238~\cite{ghissue3238}   & P1       & High     & Phantom tool execution \\
\#3376~\cite{ghissue3376}   & P2, P3   & Critical & 21\% analysis as 100\% \\
\#6049~\cite{ghissue6049}   & P2, P3   & Critical & \$200+ lost, 20+ hours \\
\#27430~\cite{ghissue27430} & P3       & Critical & Fabricated claims on 8+ platforms \\
\#32291~\cite{ghissue32291} & P4       & High     & Unfalsifiable verification \\
\#32301~\cite{ghissue32301} & P10      & High     & 5 prompts to find 4 errors \\
\#32650~\cite{ghissue32650} & P2, P10  & Critical & 30--40\% time as quality gate \\
\#33781~\cite{ghissue33781} & P3, P7   & Critical & 60k+ tokens on fake tests \\
\#35357~\cite{ghissue35357} & P8       & High     & Weaponized CLAUDE.md rules \\
\#36241~\cite{ghissue36241} & P2, P5   & High     & Found new issues each forced round \\
\#3382~\cite{ghissue3382}   & P6       & High     & 350 upvotes on sycophancy \\
\#31497~\cite{ghissue31497} & P9       & Medium   & Exploited user patience \\
\#15443~\cite{ghissue15443} & P8       & High     & Acknowledged then ignored rules \\
\#21749~\cite{ghissue21749} & P5       & Medium   & Self-aware failure admission \\
\#28828~\cite{ghissue28828} & P5, P3   & High     & Misrepresents while bug-reporting \\
\#14759~\cite{ghissue14759} & P6       & Medium   & Sycophancy undermines usefulness \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Pattern frequency analysis.}
Across 16 coded issues, 24 pattern codes were assigned (some issues exhibit multiple patterns).
The distribution is:
P3 (fabricated analysis) appears in 5 issues, making it the most frequently observed pattern.
P2 (partial work as complete) appears in 4 issues.
P5 (apology loop) appears in 3 issues.
P6 (sycophantic completion) and P8 (weaponized rules) each appear in 2 issues.
P1, P4, P7, P9, and P10 each appear in 1--2 issues.

The dominance of P3 and P2 is consistent with our root cause analysis: fabrication (P3) is driven by RC4 (shortcut exploitation), where generating a plausible description costs $\sim$50 tokens versus $\sim$5{,}000 for genuine analysis; partial completion (P2) is driven by the convergence of RC3 (context rot) and RC5 (completion pressure).
Notably, P10 (human quality gate dependency) appears in only 2 issues as an explicit code, but it is the \emph{implicit} condition for every issue in the table: each incident was detected only because a human performed verification that the model did not.


% ══════════════════════════════════════════════════════════════
\subsection{Community Workarounds}
\label{sec:workarounds}

Users have developed workarounds that fall into three categories: environment configuration, prompt engineering, and process discipline.
We catalog these not as solutions but as evidence of the problem's severity---the effort users invest in mitigation reflects the cost of the underlying failure.

\paragraph{Environment variables.}
The primary environment-level intervention is setting \texttt{CLAUDE\_CODE\_DISABLE\_ADAPTIVE\_THINKING=1}, which forces a fixed thinking budget rather than allowing the model to adaptively reduce reasoning depth.
This was confirmed by Boris Cherny of the Claude team via Hacker News~\cite{hn2026claudeissue} as a supported configuration.

\paragraph{Prompt strategies.}
Users have converged on several prompt-level interventions: using \texttt{/effort high} or the keyword \texttt{ULTRATHINK} to request maximum reasoning depth; writing 2{,}000+ word \texttt{CLAUDE.md} configuration files with explicit anti-tautological quality assurance rules; placing critical instructions at both the start and end of prompts to exploit the U-shaped attention curve documented by Liu et al.~\cite{liu2023lost}; and including explicit prohibitions against claiming completion without evidence~\cite{dolthub2025gotchas,ghissue36241}.

\paragraph{Process strategies.}
Process-level workarounds require human discipline rather than model cooperation: breaking tasks into small pieces to avoid context rot; forcing 3--4 rounds of re-analysis on each task; adopting test-driven development where humans write tests first and the model only implements; never allowing the model to manage git operations; reviewing every test change for assertion weakening; and running parallel sessions to cross-check outputs~\cite{dolthub2025gotchas}.

\paragraph{Assessment.}
Prompt strategies address symptoms: they attempt to restore reasoning depth that should be the default.
Process strategies are effective but require constant human enforcement---they are, in essence, the human quality gate that P10 describes.
Tool-based approaches, such as stop-hook enforcement~\cite{laurenzo2026amd} and adversarial code review~\cite{adversarialreviewer}, are the most effective because they operate mechanically and do not depend on model compliance.
They are also the least adopted, because they require engineering effort to implement.
This gap---between what works and what users do---is what the Completion Integrity System (Section~\ref{sec:system}) is designed to close.


% ══════════════════════════════════════════════════════════════
\subsection{Economic Impact Analysis}
\label{sec:economic}

We present cost estimates at three levels of granularity.
All estimates include explicit assumptions; readers should evaluate the conclusions in light of those assumptions.

\paragraph{Per-user quality gate cost.}
Issue \#32650 documents that users spend 30--40\% of Claude Code interaction time performing manual quality verification~\cite{ghissue32650}.
We estimate the cost of this overhead as follows.
\emph{Assumptions:} US software developer median hourly rate of \$60 (Bureau of Labor Statistics, May 2025) plus 25\% benefits overhead = \$75/hour fully loaded; median developer uses Claude Code 4 hours per day; quality gate time = 30--40\% of interaction time.
Under these assumptions: 4 hours $\times$ 0.30 = 1.2 hours/day at the low end; 4 hours $\times$ 0.40 = 1.6 hours/day at the high end.
Daily cost: \$90--\$120.
Monthly cost (20 working days): \$1,800--\$2,400.
This represents not the cost of using Claude Code, but the cost of \emph{verifying} its output---effort that produces no new value.

\paragraph{Per-incident documented costs.}
Table~\ref{tab:incident-costs} presents costs documented in specific community reports.

\begin{table}[h]
\centering
\caption{Documented per-incident costs from community reports.}
\label{tab:incident-costs}
\begin{tabular}{lll}
\toprule
Issue & Documented Cost & Category \\
\midrule
\#6049~\cite{ghissue6049}   & \$200+ in registrations, 20+ hours review & Direct financial + labor \\
\#33781~\cite{ghissue33781} & 60{,}000+ tokens ($\sim$\$40) on fabricated tests & API waste \\
\#27430~\cite{ghissue27430} & Unmeasured reputational damage, 8+ platforms & Reputational \\
\bottomrule
\end{tabular}
\end{table}

These are the incidents where users documented costs.
The selection bias is toward underreporting: most users who encounter false completions do not file detailed cost analyses.

\paragraph{API waste from behavioral degradation.}
The AMD dataset documents 80$\times$ more API requests and 64$\times$ more output tokens after thinking depth reduction, with no improvement in outcomes~\cite{laurenzo2026amd}.
If a typical pre-degradation Claude Code session costs approximately \$0.50 in API usage, the post-degradation equivalent session costs approximately \$40---for worse results.
We present this as a ratio rather than an absolute figure because per-session costs vary significantly by task complexity, model version, and pricing tier.

\paragraph{Industry-scale estimate.}
We construct a range estimate with three scenarios.
\emph{Shared assumptions:} Claude Code had approximately 500{,}000 active users as of early 2026 (based on download counts and Anthropic's public statements); not all users encounter false completion on every session; the quality gate overhead applies only during active coding sessions.

\begin{table}[h]
\centering
\caption{Industry-scale cost estimate for false completion overhead. All figures are monthly estimates with stated assumptions. These are order-of-magnitude estimates, not precise measurements.}
\label{tab:industry-cost}
\begin{tabular}{llrrr}
\toprule
Scenario & Assumption & Affected users & Monthly/user & Monthly total \\
\midrule
Low    & 10\% encounter regularly & 50{,}000  & \$900  & \$45M \\
Medium & 25\% encounter regularly & 125{,}000 & \$1{,}350 & \$169M \\
High   & 40\% encounter regularly & 200{,}000 & \$1{,}800 & \$360M \\
\bottomrule
\end{tabular}
\end{table}

We do not know which scenario is closest to reality.
The low estimate assumes that most users either do not encounter false completion frequently or do not perform systematic verification (accepting false completions unknowingly, which shifts the cost from verification overhead to defect cost).
The high estimate assumes that the patterns documented in \S\ref{sec:community-issues} are representative of a large minority of users.
In all scenarios, the monthly cost exceeds the monthly subscription revenue per affected user, suggesting that false completion imposes a net negative economic value on the user base even when the tool provides genuine productivity gains on successful interactions.

The purpose of this analysis is not to claim a precise dollar figure but to establish that the economic magnitude is significant enough to warrant systematic mitigation---the subject of Section~\ref{sec:system}.
