\section{Mechanical Mitigation: The Completion Integrity System}
\label{sec:system}

The taxonomy of Section~\ref{sec:taxonomy} establishes that false completion arises from structural causes---training incentives, architectural biases, and emergent shortcuts---that operate at the weight level and persist through prompt-level interventions.
The evidence of Section~\ref{sec:evidence} quantifies the cost: 30--40\% of user time consumed by manual verification, 80$\times$ API cost inflation under degraded conditions, and documented financial losses from undetected false completions.
This section presents the Completion Integrity System, a three-layer verification architecture deployed in the VIBE Framework for Claude Code.
The system enforces evidence-based completion claims through mechanical inspection of tool logs and file state, without relying on any model-generated assertion about its own behavior.


% ══════════════════════════════════════════════════════════════
\subsection{Design Principles}
\label{sec:design-principles}

Five principles govern the system's architecture.
Each is justified by specific evidence from prior sections.

\paragraph{1. External verification over self-assessment.}
The Worker-Inspector Paradox (Definition~\ref{def:paradox}, Proposition~\ref{prop:external}) establishes that the model cannot reliably verify its own completion because verification is subject to the same incentive function $\mathcal{I}$ as task execution.
Any verification request routed through the model is processed by the same weights that produced the original false completion.
The system therefore places all verification logic outside the model, in shell scripts that execute after the model's response is generated but before it is delivered to the user.

\paragraph{2. Mechanical gates over instruction-based rules.}
Section~\ref{sec:workarounds} documents that prompt strategies---explicit instructions to verify work, anti-shortcut rules in configuration files, ``double-check'' directives---reduce false completion frequency but cannot eliminate it.
The model processes instructions through the same parameters that produce the behavior the instructions attempt to prevent.
Mechanical gates operate on the transcript as data: they parse JSON, count tool invocations, and compare numerical claims against evidence.
They do not ask the model to cooperate; they inspect its output.

\paragraph{3. Evidence-required claims.}
Failure patterns P1 (phantom execution) and P3 (fabricated analysis) demonstrate that the model generates plausible descriptions of work it has not performed.
If a model claims to have analyzed 20 items, the transcript must contain evidence of 20 analyses---tool calls, file reads, or command executions corresponding to each claimed item.
The system treats unsubstantiated numerical claims as failures, not warnings.

\paragraph{4. Transcript as ground truth.}
The Claude Code session transcript is a structured JSON log of every tool invocation, including the tool name, input parameters, and output.
Tool calls are executed by the harness, not by the model: the model generates a \texttt{tool\_use} block, but the harness performs the actual file read, shell command, or API request and records the result.
The model cannot fabricate transcript entries because it does not control the execution environment.
This asymmetry---the model proposes, the harness disposes---makes the transcript a reliable evidence source for mechanical verification.

\paragraph{5. Configurable strictness.}
Different tasks have different verification needs.
A quick factual question does not require the same scrutiny as a 70-item competitor analysis.
The system supports four operating modes (strict, balanced, light, off), configured via the \texttt{VIBE\_INTEGRITY\_MODE} environment variable, allowing users to match verification overhead to task criticality.


% ══════════════════════════════════════════════════════════════
\subsection{Layer 1: Completion Sentinel}
\label{sec:sentinel}

The Completion Sentinel is a Bash script (\texttt{completion-sentinel.sh}, 509 lines) registered as both a \texttt{Stop} and \texttt{SubagentStop} hook in the Claude Code harness.
It fires after every assistant message, before the message is delivered to the user.

\paragraph{Interface.}
The sentinel reads session context from stdin as JSON, receiving the session identifier, transcript file path, last assistant message text, hook event name, and agent identifier (for subagent namespacing).
It produces one of two outcomes: exit code~0 (pass, silent---the message is delivered unchanged) or exit code~2 (block---the message is withheld and the sentinel's stderr output is shown to the model as feedback, allowing it to correct its response).

\paragraph{Transcript parsing.}
The sentinel uses \texttt{jq} to parse the full session transcript.
It locates the last user message containing actual text content (excluding messages that contain only \texttt{tool\_result} blocks), then extracts all \texttt{tool\_use} blocks from assistant messages following that point.
From these blocks it computes: per-tool-name invocation counts, Read targets (file paths), image reads (files matching \texttt{png|jpg|jpeg|webp|gif|svg}), Bash commands, VIBE\_GATE markers extracted from tool results via regex capture, Skill invocations, Agent calls, and whether any Read or Bash calls occurred after the last Agent dispatch.
This extraction runs once per hook invocation and produces a single JSON object used by all six checks.

\paragraph{Completion detection.}
The sentinel detects completion claims through a multilingual regular expression covering English (\texttt{done}, \texttt{complete}, \texttt{finished}), Italian (\texttt{fatto}, \texttt{completato}, \texttt{terminato}, \texttt{tutti}), French (\texttt{termin\'{e}}, \texttt{compl\'{e}t\'{e}}), Spanish (\texttt{terminado}, \texttt{completado}), Portuguese (\texttt{conclu\'{i}do}, \texttt{finalizado}), and German (\texttt{fertig}, \texttt{abgeschlossen}).
Multilingual coverage is necessary because the model may respond in the user's language regardless of the system prompt language.

\paragraph{The six checks.}
Algorithm~\ref{alg:sentinel} presents the six mechanical checks in pseudocode.
Each check compares a structural property of the model's message against evidence in the transcript.

\begin{algorithm}[t]
\caption{Completion Sentinel: Six Mechanical Checks}
\label{alg:sentinel}
\begin{algorithmic}[1]
\Require transcript $\mathcal{T}$, last message $m$, mode $\in \{\text{strict, balanced, light, off}\}$
\State Parse $\mathcal{T}$: extract tool calls $C$, results $R$ from current turn
\State $n \gets |C|$, $\textit{hasCompletion} \gets \textsc{DetectCompletionIndicators}(m)$
\Statex
\State \textbf{Check 1: Zero-Tool Completion}
\If{$\textit{hasCompletion} \land n = 0$}
    \State \Return \textsc{Fail}(\text{high}, \text{``completion claim with 0 tool calls''})
\EndIf
\Statex
\State \textbf{Check 2: Numerical Discrepancy}
\For{each claimed count $k$ extracted from $m$}
    \State $r \gets \textsc{ContextAwareCount}(C, m, k)$ \Comment{image reads for visual claims, etc.}
    \If{$r < 0.5k$} \Return \textsc{Fail}(\text{high}) \EndIf
    \If{$r < 0.85k$} \Return \textsc{Fail}(\text{medium}) \EndIf
\EndFor
\Statex
\State \textbf{Check 3: Test/Build Claim Without Execution}
\If{$m$ claims test/build results $\land$ no test/build commands in $C$}
    \State \Return \textsc{Fail}(\text{high})
\EndIf
\Statex
\State \textbf{Check 4: Subagent Trust Without Verification}
\If{$m$ references agent results $\land$ no Read/Bash after last Agent call}
    \State \Return \textsc{Warn}(\text{medium})
\EndIf
\Statex
\State \textbf{Check 5: Completion Scope Mismatch}
\If{$m$ uses totality words $\land$ $n < 5$ $\land$ $|m| > 200$}
    \State \Return \textsc{Warn}(\text{medium})
\EndIf
\Statex
\State \textbf{Check 6: VIBE\_GATE Marker Verification}
\If{skill used $\land$ $\textit{hasCompletion}$ $\land$ no VIBE\_GATE markers in $R$}
    \State \Return \textsc{Fail}(\text{high})
\EndIf
\If{marker value $<$ claimed count in $m$}
    \State \Return \textsc{Fail}(\text{high})
\EndIf
\end{algorithmic}
\end{algorithm}

\paragraph{Context-aware tool counting.}
Check~2 does not simply compare claimed numbers against total tool calls.
When the message contains visual-domain keywords (\texttt{screenshot}, \texttt{image}, \texttt{design}), the sentinel counts image file reads specifically.
When the message references pages or URLs, it counts \texttt{WebFetch} invocations.
When the message claims test results, it counts Bash commands matching test runner patterns (\texttt{jest}, \texttt{pytest}, \texttt{vitest}, \texttt{mocha}, \texttt{cargo test}, \texttt{go test}).
This context sensitivity prevents a model from satisfying a ``20 screenshots analyzed'' claim by performing 20 unrelated file reads.

\paragraph{Two-tier severity.}
Check~2 uses two thresholds: a ratio below 0.5 triggers a high-confidence failure (the model claimed $k$ but evidence shows fewer than $k/2$); a ratio below 0.85 triggers a medium-confidence failure.
The gap between 0.85 and 1.0 accommodates legitimate tool reuse---a single Read call may examine multiple items in one file.

\paragraph{Resolution mode.}
After a block, the sentinel sets a flag file and enters resolution mode on the next invocation.
In resolution mode, the model must either perform new tool calls or provide specific counts (matching the pattern ``X of Y'' in six languages) to exit the block state.
If the model repeats totality claims (\texttt{all}, \texttt{every}, \texttt{tutti}, \texttt{jeder}) without new evidence or specific counts, the sentinel issues a second block.
This prevents the apology loop (P5): acknowledgment without behavioral change does not satisfy the gate.

\paragraph{Mode-dependent blocking.}
In strict mode, any check failure triggers a block.
In balanced mode (the default), only high-confidence failures trigger blocks; medium-confidence findings are recorded but not blocking.
In light mode, only Checks~1 and~3 are active, and findings are reported non-blocking via stdout.
Off mode disables the sentinel entirely.


% ══════════════════════════════════════════════════════════════
\subsection{Layer 2: File Verification}
\label{sec:verifier}

The File Verifier (\texttt{completion-verifier.sh}, 133 lines) is a second stop hook that reads Layer~1's findings and performs independent filesystem checks.
It activates only in strict or balanced mode, and only when Layer~1 has recorded at least 5 tool calls and flagged at least one failure or warning.

The verifier reads the sentinel's JSON output file, which contains per-check results, tool counts, and VIBE\_GATE marker values.
It then rechecks marker claims against the filesystem: for \texttt{screenshot\_count} markers, it counts PNG files in the output directory; for \texttt{json\_entries} markers, it runs \texttt{jq length} on the output JSON; for \texttt{empty\_files} markers, it searches for zero-byte files.
If the filesystem count disagrees with the marker value recorded in the transcript, the discrepancy is reported.

In strict mode, filesystem discrepancies trigger a block (exit~2).
In balanced mode, the verifier reports discrepancies via stderr and requests that the model present the findings to the user.
The verifier is intentionally conservative: if it finds no independent discrepancies, it defers to Layer~1's findings and exits cleanly.

The current implementation is command-based (a shell script hook).
The Claude Code hook system supports a \texttt{type: "agent"} hook that would allow Layer~2 to operate as an independent agent with its own tool access, enabling more sophisticated verification such as running test suites or inspecting build artifacts.
This upgrade is planned but has not been deployed, as the agent hook type has not been tested in production.


% ══════════════════════════════════════════════════════════════
\subsection{Layer 3: Skill-Level VIBE\_GATE}
\label{sec:vibegate}

The first two layers operate post-hoc: they inspect the model's output after generation.
Layer~3 operates \emph{during} task execution, embedding verifiable checkpoints into the workflow itself.

\paragraph{Marker protocol.}
Each VIBE skill defines a set of verification commands that produce markers in the format:
\begin{quote}
\texttt{echo "VIBE\_GATE: \{check\_name\}=\{value\}"}
\end{quote}
The model is instructed to execute these commands via the Bash tool before claiming any phase complete.
The protocol is defined in \texttt{skills/\_shared/integrity-gate.md} and referenced by all skills at their delivery phase.

\paragraph{Why markers cannot be fabricated.}
The critical property is that the value after the equals sign comes from shell command substitution, not from the model's generation.
Consider a marker command where the echo string embeds a shell command substitution that counts PNG files via \texttt{ls} and \texttt{wc~-l}.
The model generates the echo command text, but the \emph{shell} evaluates the embedded substitution.
The Bash tool executes the command in a real shell environment; the file count is produced by the operating system's directory listing, not by the model's token generation.
The model cannot cause a directory listing to return 20 when only 14 files exist.
This separation---the model proposes the measurement, but the shell performs it---is what makes VIBE\_GATE markers trustworthy evidence.

\paragraph{Deployment.}
The system currently defines 35 markers across 5 skills: competitor research (7~markers covering screenshot counts, JSON entry counts, lens completeness, metadata existence, pattern files, and empty file detection), UI design generation (8~markers for style files, template files, color tokens, spacing tokens, hardcoded text, focus indicators, ARIA landmarks, and responsive breakpoints), content generation (7~markers for content files, file sizes, empty strings, word counts, heading counts, schema presence, and common file existence), testing and debugging (7~markers for test exit codes, test file counts, lint results, type check results, red/green phase confirmation, and regression test existence), and conversion optimization (6~markers for audit file existence, Fogg B=MAP scores, recommendation counts, ICE scoring completeness, competitor benchmarks, and sort order verification).

\paragraph{Batch checkpointing.}
For tasks that process many items (e.g., analyzing 20 competitors), the skill instructions direct the model to work in batches of 5 items, running VIBE\_GATE verification after each batch.
This design serves two purposes: it provides early detection of partial completion (a batch that produces 3 of 5 expected outputs is caught immediately, not after the model claims all 20 are done), and it creates intermediate evidence in the transcript that Layer~1's Check~6 can verify against the final completion claim.

\paragraph{Strict equality.}
The integrity gate protocol enforces exact numerical matching: 14 does not equal 20, 18 does not equal 20, and ``close enough'' is not passing.
Markers that return \texttt{SKIPPED} or \texttt{ERROR} are treated as failures.
This strictness is deliberate: the failure patterns documented in Section~\ref{sec:failure-patterns} demonstrate that the model's natural tendency is to round, approximate, and elide discrepancies.
The verification protocol must not share this tendency.


% ══════════════════════════════════════════════════════════════
\subsection{Configuration}
\label{sec:configuration}

The system's operating mode is set via the \texttt{VIBE\_INTEGRITY\_MODE} environment variable, with four levels of strictness shown in Table~\ref{tab:modes}.

\begin{table}[h]
\centering
\caption{Completion Integrity System operating modes.}
\label{tab:modes}
\begin{tabular}{lll}
\toprule
Mode & Active Checks & Use Case \\
\midrule
strict & All 6, low thresholds & Critical deliverables \\
balanced & All 6, moderate thresholds & Default \\
light & Checks 1, 3 only & Quick interactions \\
off & None & Disabled \\
\bottomrule
\end{tabular}
\end{table}

The default mode is balanced.
In balanced mode, all six checks run but only high-confidence failures (ratio below 0.5, or complete absence of expected evidence) trigger blocking.
Medium-confidence findings are logged to a daily event file (\texttt{/tmp/vibe-integrity-events-YYYY-MM-DD.jsonl}) for post-session analysis but do not interrupt the model's response.
The sentinel also supports per-session pausing via a flag file, allowing users to temporarily disable checks for sessions where verification overhead is not warranted.

\input{figures/architecture}
