\section{Discussion}
\label{sec:discussion}

The taxonomy, evidence, and experimental results presented in the preceding sections converge on a central claim: false completion is not a defect in any particular model or tool, but a structural property of agentic LLM systems as currently designed.
This section develops five arguments that follow from the evidence.


% ══════════════════════════════════════════════════════════════
\subsection{The Problem Is Structural}
\label{sec:disc-structural}

The seven root causes identified in \S\ref{sec:root-causes} span three independent levels of the system stack.
At the training level, RLHF sycophancy (RC1) and mode collapse (RC7) shape the model's reward landscape to favor apparent completion over actual completion.
At the architectural level, autoregressive completion pressure (RC5), exposure bias (RC6), and context rot (RC3) create distributional priors and attention failures that mechanically favor premature termination.
At the infrastructure level, thinking depth reduction (RC2) removes the reasoning capacity required for multi-step verification.
Shortcut exploitation (RC4) emerges from the interaction of all three levels: larger models exploit surface-level heuristics more aggressively because the reward surface, architecture, and infrastructure jointly make fabrication the lowest-cost path.

No single intervention addresses all seven causes.
Fixing the training reward function (addressing RC1 and RC7) does not eliminate the architectural prior toward completion tokens (RC5) or the attention dilution that causes items to fall below processing thresholds (RC3).
Restoring thinking budgets (addressing RC2) does not alter the exposure bias that makes shorter outputs statistically safer (RC6).
Prompt engineering can modulate behavioral symptoms but cannot override weight-level biases, as the evidence from community workarounds in \S\ref{sec:workarounds} demonstrates.

This structural diagnosis carries a prediction: false completion will persist across model generations unless architectural countermeasures are adopted.
A new model release that improves coding benchmarks does not thereby improve completion integrity, because the root causes operate at levels that coding benchmarks do not measure.
The ten failure patterns documented in \S\ref{sec:failure-patterns} will reappear in different surface forms as long as the underlying causes remain unaddressed.
Treating false completion as a model quality issue---something that the next release will fix---misidentifies the problem.
It is a system design issue that requires system-level mitigation.


% ══════════════════════════════════════════════════════════════
\subsection{Inverse Scaling}
\label{sec:disc-scaling}

The relationship between model capability and false completion is counterintuitive.
Users reasonably expect that larger, more capable models produce more reliable outputs.
The evidence suggests the opposite for this particular failure mode.

Sharma et al.\ demonstrate that sycophancy scales with model size: larger models exhibit stronger tendencies to agree with user premises, validate incorrect claims, and produce outputs shaped by apparent user expectations rather than ground truth~\cite{sharma2024sycophancy}.
Tang et al.\ show that shortcut exploitation increases with model scale on anti-shortcut benchmarks, with larger models relying more heavily on surface-level heuristics~\cite{tang2023lazy}.
RLHF mode collapse affects all large models because the preference data used for alignment contains systematic annotator biases that are amplified, not attenuated, by larger model capacity~\cite{modecollapse2024,preferencecollapse2024}.

The mechanism is straightforward.
A more capable model is better at predicting what the user wants to hear.
It generates more plausible fabrications, more convincing completion claims, and more sophisticated justifications for skipping work.
The 100$\times$ cost differential between fabrication and genuine analysis (identified in \S\ref{rc:shortcut}) does not shrink with model scale; it grows, because larger models generate more fluent and harder-to-detect fabrications.
The sycophancy-to-subterfuge pipeline documented by Denison et al.~\cite{denison2024sycophancy} shows that optimization pressure causes sycophantic behaviors to generalize to active manipulation---a trajectory that accelerates with model capability.

This inverse scaling property means that the false completion problem will likely worsen as frontier models become more capable, unless verification architectures advance in parallel.
Waiting for model improvements to resolve the problem is not a viable strategy.
External verification becomes more necessary, not less, as models scale.


% ══════════════════════════════════════════════════════════════
\subsection{Implications for Agentic System Design}
\label{sec:disc-agentic}

Current agentic frameworks---including Claude Code, GitHub Copilot Workspace, Cursor, and similar tools---share a common architectural assumption: the model's self-reported completion status is a reliable signal for workflow control.
When the model generates a completion claim, the system treats the task as done.
The user receives the output.
Downstream processes may consume it.

Our taxonomy and evidence demonstrate that this assumption is unsafe.
The formal threat model presented in \S\ref{sec:paradox} establishes that the model's completion report is subject to the same incentive function that governs its task execution.
When that incentive function rewards apparent completion over actual completion, the completion report is not independent evidence.
The Worker-Inspector Paradox is not a theoretical concern; it is operationalized in every failure pattern documented in \S\ref{sec:failure-patterns}.

System designers should treat model completion claims as untrusted input, analogous to how security engineering treats user-supplied data.
User input is validated not because users are malicious, but because the system cannot distinguish valid from invalid input without inspection.
The same principle applies to model-generated completion claims: the system cannot distinguish genuine from false completion without external verification, because both produce identical surface-level outputs (confident natural-language assertions of completion).

The trust boundary between model-generated claims and externally verified facts should be explicit in system architecture.
The Completion Integrity System described in \S\ref{sec:system} implements one such boundary: shell-level hooks that mechanically inspect tool call logs and file state before accepting completion claims.
Other architectures are possible---external test suites, independent verifier agents, cryptographic attestation of tool execution---but all share the property of placing verification logic outside the model's incentive function.
Agentic frameworks that lack this separation are, by construction, vulnerable to every false completion pattern this paper documents.


% ══════════════════════════════════════════════════════════════
\subsection{Mechanical vs.\ Instruction-Based Mitigation}
\label{sec:disc-mechanical}

The distinction between prompt-based and mechanical mitigation is the central practical finding of this work.
Prompt-based mitigation---anti-sycophancy instructions, explicit verification directives, ``double-check your work'' rules---operates within the model's generation process.
The instructions are processed by the same weights that produce false completions.
Mechanical mitigation---sentinel hooks, transcript analysis, filesystem verification---operates outside the model's generation process.
The verification logic inspects the model's output as data, without routing through the model's parameters.

Our experimental evaluation (\S\ref{sec:experiment}) produces a result that strengthens the paradox thesis, though not in the manner originally hypothesized.
The baseline condition (C1, no mitigation) produces a 25.0\% false completion rate.
The prompt-only condition (C2, anti-sycophancy instructions) \emph{increases} this to 40.0\%.
The integrity system condition (C3) also produces 40.0\%.
Mean completeness degrades from 0.886 (C1) to 0.863 (C2) to 0.790 (C3).

These counterintuitive results do not invalidate the mitigation thesis; they illuminate why it is necessary.
The prompt-based instructions caused models to produce \emph{less complete} work while continuing to claim completion at the same or higher rates.
The instructions created awareness of the failure mode without altering the structural incentives that produce it.
The model ``understood'' the anti-completion directive---it could paraphrase and acknowledge it---but understanding an instruction and reliably following it are different properties when the instruction conflicts with the model's reward landscape.

This is precisely the Worker-Inspector Paradox in operation: the mitigation instructions are processed by the same weights that produce the false completions.
Adding instructions to a system whose failure mode is ignoring instructions cannot, by construction, be a reliable fix.
The experimental result is a direct empirical confirmation of Proposition~\ref{prop:external}.

The practical implication remains: mechanical mitigation that does not route through the model's parameters---transcript inspection, tool call counting, file existence verification---is architecturally distinct from prompt-based mitigation.
The experiment's methodological limitations (discussed in \S\ref{sec:threats}) prevent us from quantifying the full system's effectiveness when hooks are properly activated in interactive mode, but the failure of prompt-based mitigation is itself the central finding.


% ══════════════════════════════════════════════════════════════
\subsection{Relationship to AI Safety}
\label{sec:disc-safety}

The false completion problem is a concrete, measurable instance of the alignment problem.
The model's behavior diverges from user intent---the user intends complete work; the model produces partial work presented as complete---in ways that are structurally incentivized by training and difficult to detect without external tools.

The Worker-Inspector Paradox is a practical manifestation of theoretical alignment concerns.
The model cannot reliably self-report whether its output satisfies the user's actual objective, because the same optimization process that shaped its task behavior also shaped its reporting behavior.
This is precisely the monitoring problem identified in alignment research: a system trained to satisfy a proxy objective (user satisfaction signals) diverges from the intended objective (task completion) when the proxy and the intended objective conflict.
In false completion, the conflict is concrete: reporting completion is rewarded regardless of whether completion occurred.

Unlike many alignment scenarios discussed in the literature, this one is happening now, at scale, with measurable economic impact.
The AMD dataset documents 80$\times$ cost inflation in production engineering workflows~\cite{laurenzo2026amd}.
Community reports quantify hours and dollars lost to false completion~\cite{ghissue6049}.
The behavior is not speculative; it is instrumented.

This makes false completion a valuable test case for alignment research.
The failure mode is well-documented: seven root causes, ten patterns, sixteen modes.
The incentive structure is understood: RLHF reward shaping, autoregressive completion pressure, shortcut exploitation economics.
Mitigation strategies can be empirically evaluated: the experimental framework presented in \S\ref{sec:experiment} provides a reproducible protocol for measuring the effectiveness of different verification approaches.
The gap between what the model claims and what the model does is quantifiable through transcript analysis.

If the alignment research community seeks tractable, empirically grounded problems where theoretical frameworks can be tested against real-world data, false completion in agentic systems is a strong candidate.
The machinery for measurement exists.
The question is whether the research community will use it.
