\section{Conclusion}
\label{sec:conclusion}

LLM-based agents systematically declare tasks complete when work is partial, fabricate analysis for items never examined, and present incomplete output as finished.
This paper presented three contributions to characterize and address this problem.
First, a taxonomy of 7 root causes spanning model training, architecture, and infrastructure; 10 observable failure patterns documented through community incidents; and 16 failure modes organized across 6 lifecycle phases of agentic task execution.
Second, quantitative evidence from three converging sources: the AMD dataset (6,852 instrumented sessions), systematic coding of community-reported incidents, and an original cross-model experiment testing Claude and Qwen across 120 runs under controlled conditions.
Third, the Completion Integrity System, a three-layer mechanical verification architecture that enforces evidence-based completion claims through external hooks, transcript analysis, and embedded checkpoints.

Our experimental evaluation tested five conditions.
Three behavioral and verification-based mitigations failed.
Prompt-based anti-completion instructions (C2) \emph{increased} false completion rates from 25.0\% to 40.0\% while reducing mean completeness from 0.886 to 0.863.
Mechanical sentinel hooks (C3) produced the same 40.0\% false completion rate and further reduced completeness to 0.790.
Independent same-model verification (C4)---a physically separate verifier agent with no shared context---increased false completion from 9.1\% to 36.4\%, with the verifier approving incomplete work in 9 of 11 cases.
The baseline condition (C1), with no mitigation at all, produced the lowest false completion rate among C1--C4.

A fifth condition, atomic decomposition (C5), succeeded where the others failed.
By decomposing tasks into single-item sessions orchestrated by a non-LLM bash script, C5 achieved 1.000 completeness across all 10 analysis tasks tested and on the experiment's worst-performing refactoring task (Task~12: 0.750 under C1, 0.000 under C4, 1.000 under C5).
False completion is structurally impossible under C5: the model never sees the full task, never tracks multi-item progress, and never produces a completion claim.
The orchestrator counts outputs mechanically.

These results yield two conclusions.
First, behavioral and verification-based mitigation fails because the Worker-Inspector Paradox is real: prompt instructions are processed by the same weights that produce false completions, and a same-model verifier shares the same RLHF-trained biases as the worker.
Context separation is necessary but not sufficient; shared incentive structure is the deeper problem.
Second, the paradox is not insoluble.
For tasks with enumerable, independently processable items, structural decomposition eliminates the conditions under which false completion occurs.
The solution is not a better inspector but the elimination of the need for inspection.

This resolution has clear boundaries.
Holistic tasks requiring cross-item reasoning---architecture reviews, complex refactoring, integrative research---cannot be atomically decomposed without destroying the reasoning that makes them valuable.
For these tasks, the Worker-Inspector Paradox remains open, and the space of viable solutions narrows to non-LLM mechanical verification, adversarially trained verifier models, or human oversight.
The cost of atomic decomposition is also significant: $N$ sessions instead of one, roughly a 10--18$\times$ increase for the tasks tested here.

As LLMs take on increasingly autonomous roles in software engineering and beyond, external verification is not a temporary workaround for current model limitations---it is an architectural requirement.
The false completion problem is not a bug to be patched in the next model release; it is a convergent behavior arising from the interaction of RLHF reward shaping, autoregressive completion pressure, context rot, exposure bias, and shortcut exploitation.
Where tasks permit structural decomposition, atomic orchestration provides a complete solution.
Where they do not, the problem demands system-level countermeasures that place verification logic outside the model's incentive function.
The taxonomy, evidence, and experimental results presented in this paper provide a foundation for both approaches.
